---
title: "HW2-EDA with R - Part1: Plotting"
author: "misken"
date: "September 29, 2019"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset of US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean. The official documentation for the ACS data can be found at
https://www2.census.gov/programs-surveys/acs/tech_docs/subject_definitions/2017_ACSSubjectDefinitions.pdf.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

**Hacker Extra** - Put this project folder under version control using git. Only
track the .R file, the .Rmd file and your .gitignore file. Create a remote repo
for this homework assignment in your GitHub account. Periodically push your commits
to GitHub. When done with the assignment, paste in the URL of your GitHub project
so that I can go look at it.

## Step 3 - Explore data folder

Inside of the main project folder you'll find the data files for this assignment in a subfolder called `data/`.

- **acs2017_census_tract_data.csv**
- **acs2017_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data. **You do NOT need to rerun the census_data_prep.R file.** I'm
including it and encouraging you to take a look through it to get a sense of what a
typical data prep script might look like. We'll be learning to do all of the things
shown in that file.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.

Note that if you have errors in your R code, **knitr** will stop at the error
and won't create an html file. Sooo, you should comment out R code that has
errors and write a little comment stating that it's an error. Keep doing this
until you successfully knit an html file based on your R Markdown document.
You'll be submitting both the html file as well as the Rmd file.

You'll be submitting BOTH the completed R Markdown file as well as the generated HTML file. You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(ggplot2)
```


### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_mi_df.csv** into a data frame named county_mi_df.

```{r read_countydata}
county_df <- read.csv(file = "./data/county.csv", header = TRUE)
county_mi_df <- read.csv(file = "./data/county_MI.csv", header = TRUE)

```

* What does each row represent?

Each Row reperesents a county and its population information 

Check the structure of each data.frame and use the `summary` function to begin
exploring the data. 

* Are there any NA values? 
* What differences do you notice between the structure of the two data frames.


```{r}
str(county_df)
str(county_mi_df)
summary(county_df)
summary(county_mi_df)
```

There seems to be no NA values present inside the data

The two dataframes seem to be identicle in structure

**Unless otherwise specified, you should use the ggplot2 package for all of the
following plots.**

### Problem 2 - Basic plots for Income

Create a histogram and a kernel density plot for Income and for IncomePerCap for the entire US (i.e. use the `county_df` dataframe.) The Income field is median income for the county and IncomePerCap is aggregate income divided by population size, for the county.

In addition, compute and display the following statistics for Income and IncomePerCap:

* mean and median
* standard deviation
* min, max and range
* 95th and 99th percentile
* IQR
* coefficient of variation


```{r summary_stats}
mean(county_df$Income)
median(county_df$Income)
summary(county_df$Income)
```

```{r histo_dens}
ggplot(data=county_df, aes(x = Income)) + 
  geom_histogram()

ggplot(data=county_df, aes(x = IncomePerCap)) + 
  geom_histogram()
```


Write a short paragraph describing the distribution of of these two variables.

The Income per cap has a much tigher bell curve than just average income 
>Write more about the 2 variables

Repeat the histogram for Income but make the following style changes to the plot.

* Add a title such as "Distribution of median income for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Median income"
* Change the colors of the bars to something other than the default

```{r histo_stylechanges}

ggplot(data = county_df) + 
  geom_histogram(aes(x = Income), fill="steelblue", colour="Black") +
  labs(x = "Median income", y = "Number of counties", title = "Distribution of median income for US counties") 

```

Now create a set of histograms for Income in which you facet by Division.

```{r histos_division}
 ggplot(county_df) + 
   geom_histogram(aes(x=Income)) +
   labs(x = "Median income", y = "Number of counties", title = "Distribution of median income for US Divisions") +
    facet_grid(. ~ Division)
```

Since the number of counties varies by Division, create a new version of the plot above but modify so that the y-axis shows frequency percentages instead of
counts. Hint: `..density..`

```{r histos_division_density}
 ggplot(county_df) + 
   geom_histogram(aes(x=(Income))) +
   facet_grid(. ~ Division) +
   scale_y_continuous(labels = scales::percent) + 
  labs(x = "Median income", y = "Percent of counties", title = "Distribution of median income for US Regions") 
```


Now create boxplots and violin plots for the same two variables. However, you'll
see that for both of these plot types, ggplot wants an `x` aesthetic to act as a grouping variable. Use the `region` field as your `x` aesthetic. Note anything of interest revealed by these plots.

```{r box_violin}
# Boxplots and violin plots
ggplot(county_df, aes(x = Region, y = Income)) +
  geom_boxplot() +
  labs(x = "Region", y = "Median Income", title = "Boxplot of median income for US Regions") 

ggplot(county_df, aes(x = Region, y = Income)) +
  geom_violin() + 
  labs(x = "Region", y = "Median Income", title = "ViolinPlot of median income for US Regions")

```

what these graphs reveal is that all regions but the souths largest percentage of median income is around 50,000 a year but what is also intresting is that the south has the lowest median by the largest margine but also the largest median out of any other region giving it the largest variation. However when looking at the boxplot it is revealed that the south has by far the lowest avearage and quartiles and all of the values thtat were show as the highest in the violin plot are all outliers.

Repeat the histogram of Income but only for counties with less than 500,000 people. Do NOT use
dplyr to do the filtering - instead, use boolean indexing (see intro to dataframes Rmd file).

```{r histo_lt1.5m1}
#isLessThan500K <- county_df[county_df$TotalPop <  500000,]

isLessThan500K <- county_df$TotalPop <  500000

ggplot(data=county_df[isLessThan500K,], aes(x = Income)) + 
  geom_histogram() + 
   labs(x = "Median Income", y = "Count", title = "Count of Average income for counties with less than 500,000 people")


```

**Hacker Extra** - Also create a kernel density plot of Income overlaid on a histogram of
Income. 

```{r hackerxtra_kdoverlay}

ggplot(data=county_df, aes(x = Income)) + 
  geom_histogram() +
  geom_density(color="blue")

```

### Problem 3 - How do Michigan's peninsulas compare?

Create violin plots showing Income by peninsula. Make sure
your graph has a title and good axis labels.

```{r vp_inc_by_peninsula}

ggplot(county_mi_df, aes(x = '' , y = Income)) +
  geom_violin() + 
  facet_grid(. ~ peninsula) + 
   labs(x = "", y = "Median Income", title = "Violin Plot of median income for Upper and Lower Peninsula")
  

```

Do the peninsulas 
differ with respect to distribution of median income over the counties?

The Graph above shows that the Lower Peninsula has a much wider variation of average income. In the lower Peninsula the lowest average is lower than the Uppers lowest and also the Lowers highest income is significantly larger than the highest that we see in the upper.

Create several boxplots and/or violin plots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

```{r boxplots_upvlp}
ggplot(county_mi_df, aes(x = '' , y = Income)) +
  geom_violin() + 
  facet_grid(. ~ peninsula) + 
   labs(x = "", y = "Median Income", title = "Median income for Upper and Lower Peninsula")
  
ggplot(county_mi_df, aes(x = '' , y = TotalPop)) +
  geom_violin() + 
  facet_grid(. ~ peninsula) +
  labs(x = "", y = "Total Population", title = "Total County Population for Upper and Lower Peninsula")
  
ggplot(county_mi_df, aes(x = '' , y = Transit)) +
  geom_violin() + 
  facet_grid(. ~ peninsula) + 
  labs(x = "", y = "Public Transit Usage Percentage", title = "Public Transit Usage for Upper and Lower Peninsula")
  
ggplot(county_mi_df, aes(x = '' , y = MeanCommute)) +
  geom_violin() + 
  facet_grid(. ~ peninsula) + 
  labs(x = "", y = "Average Commute", title = "Average Commute for Upper and Lower Peninsula")
  

ggplot(county_mi_df, aes(x = '' , y = Unemployment)) +
  geom_violin() + 
  facet_grid(. ~ peninsula) + 
  labs(x = "", y = "Unemployment Percentage", title = "Unemployment percentage for Upper and Lower Peninsula")
  
```

As shown in the charts above the Lower Peninsula having a higher average income is no suprise due to the populaton density between them which is shown in the second graph. Although what the second graph shows us is that a large ammount of the area in the lower peninsula is like the upper with a low population density but the lower peninsula has some areas with a very large amount of population. Looking at the third set we look at percentage of people that use public transportation to get to work. this is intresting to look at because a resonable assumtion would be that a higher percentage of people use public transportation in the lower peninsula than the upper but based off the graph it seems to be hard to tell. In the lower we have a small ammount of area in the higher percentages but a vast majority tapers off avbove 1%. In the upper penisula it seems to be more intresting in that like the lower it tapers off close to the 1% line but then balloons out from 1.5 to just over 2%. this shows that although the upper peninsula is though to be just small towns it seems that the public transportation is used quite a but. The fourth chart  shows mean communte for upper bs lower peninsula. this shows that in the lower peninsula our mean commute is significantly higher than someone in the upper peninsula. The highest average commute for someone living in the UP is just unfer 25 min with a minumum below 15 however in the LP we have a high that is approching 35 min and a low around 17 showing that people living in the lower peninsula on average have a much higher average commute. Moving to the final graph we look at unemployment rates between the upper and lower peninsula. when looking at the graph we can see that the UP has a higer unemployment rate than the lower considering the lowest for the LP is less than 5% while the lowest for the up is just under 6% and for the high end the highest in the LP is just over 12 while for the UP it is over 13%. The intresting thing that is shown in the graph however is that both seem to have the mean of the unemployment rate hovering just above 6%.


### Problem 4 - Basic bar chart

Create a bar chart showing the number of counties by State in the US. Make sure
your graph has a title and good axis labels. Figure out how to format the
x-axis labels so that they don't obscure each other.

```{r bar_counties_by_state}

ggplot(county_df, aes(x = State)) +
  geom_bar(aes(color = Region)) + 
  scale_x_discrete(expand = waiver()) + 
  labs(x = "State Name", y = "Number of Couties", title = "Number of Counties by State")
  

```

**Hacker Extra** Modify the above bar plot to be horizontal and have bars sorted from longest to shortest. In addition, map the bar color to the Region variable.

```{r bar_counties_by_state_hackerxtra, fig.height=6}


```

### Problem 5 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
IncomePerCap and MeanCommute.

```{r xy_inc_commute}

ggplot(county_mi_df, aes(x = IncomePerCap , y = MeanCommute)) +
 geom_point() + 
  labs(x = "Income Per Capita", y = "Average Commute", title = "Average Commute vs. Income per capita for Michigan")
  

```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula and the size is mapped to total population. In addition,
create a similar plot comparing Income and Professional.

```{r xy_inc_commute_peninsula}
ggplot(county_mi_df, aes(x = IncomePerCap , y = MeanCommute)) +
 geom_point(aes(size = TotalPop, color = peninsula)) + 
  labs(x = "Income Per Capita", y = "Average Commute", title = "Average Commute vs. Income per capita for Upper & Lower peninsula")
  

ggplot(county_mi_df, aes(x = IncomePerCap , y = Professional)) +
 geom_point(aes(size = TotalPop, color = peninsula)) +
  labs(x = "Income Per Capita", y = "Professional Percentage", title = "Income per capita vs. Professional Percentage for Upper & Lower peninsula")
```

The first graph shows that the mean commute seems to not have a large affect on income but rather that the upper vs lower peninsula has the larger affect on the average income per capita. When you add the color to show the Upper and Lower peninsula it is very easy to seee that all of the UP points are focused in the lower left of the graph, meaning that they have shorter commutes but lower income. This graph back up my discoveries earlier when I came to the same conclusion in the violin plots. When looking at the second graph there are a couple conclusions that could be drawn from a quick analysis. First, for the LP data it is clear to see that there is a linear positive relationship between professional population and income per capita meaning that the more percentage of perfesionals in a county the more likely the income per capita is tio be larger. The same relationship appears to be developing for the UP also but not nearly as strong as the LP. additionaly there is one large outlier where almost 40% of the population is considered professional but the income per capita does not match with the rest of the UP datapoints. This also shows that there may be a relationship between population and professionals or Income per capita but it would need to be analyzed further to truly determin if there is a realationship.

Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?


```{r xy_faceted_peninsula1}

ggplot(county_mi_df, aes(x = IncomePerCap , y = Professional)) +
 geom_point(aes(size = TotalPop)) +
  facet_grid(. ~ peninsula) + 
  labs(x = "Income Per Capita", y = "Professional Percentage", title = "Income per capita vs. Professional Percentage for Upper & Lower peninsula")

ggplot(county_mi_df, aes(x = IncomePerCap , y = Poverty)) +
 geom_point(aes(size = TotalPop)) +
  facet_grid(. ~ peninsula) + 
  labs(x = "Income Per Capita", y = "Poverty Percentage", title = "Income per capita vs. Poverty Percentage for Upper & Lower peninsula")

```

Now use the full US data frame, county_df.

Create a scatter plot of TotalPop by IncomePerCap.

```{r xy_pop_inc}

ggplot(county_df, aes(x = IncomePerCap , y = TotalPop)) +
 geom_point() +
  labs(x = "Income Per Capita", y = "Total Population", title = "Income per capita vs. Total Population for united States")

```


The widely varying values make this difficult to interpret in terms of looking
for patterns. So, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
Write a short paragraph on why one might want to use a log scale or log
transform of data when plotting.


```{r log10}

ggplot(county_df, aes(x = IncomePerCap , y = log10(TotalPop))) +
 geom_point() + 
  labs(x = "Income Per Capita", y = "Log 10 of Total Population", title = "Income per capita vs. Log 10 of Total Population for united States")

ggplot(county_df, aes(x = IncomePerCap , y = log10(TotalPop))) +
 geom_point(aes(color = Region)) + 
  labs(x = "Income Per Capita", y = "Log 10 of Total Population", title = "Income per capita vs. Log 10 of Total Population for united States")

```

Using log10 for the population variable allows us to better visualize the relationship between total populaton and income per capita. The reason this helps is that when looking at the origional graph above you can see that most of the datapoints fall near the bottom of the scale but a small number of outliers cause the scaling of the graph to strech to be able to accomodate them. This causes issues when viewing the graph because you are not concerned as much with outliers but more with the other infomation which becomes very hard to read. However when you use the log of a variable on the graph it works to correct this issue by brining the outliers in and making the graph more readable. 

## Deliverables

Make sure you knit your Rmd file. As discussed earlier, if you have errors, your document won't knit. Comment out error lines and put a comment in saying that you couldn't get the line to work. After your file is knitted into an html file, simply create a compressed archive (either tar.gz or .zip) of your project folder and upload the compressed archive file into Moodle.